---
title: "EPC data cleaning"
output: html_document
date: "2025-11-27"
---

# 1. Data downloads

```{r}
library(data.table)
library(here)
library(stringr)
library(sf)
library(tmap)
library(tidyverse)
library(janitor)
library(scales)
```

## 1.1 32 LAs EPC –\> 1 London EPC

1.  Load EPC csv files for each local authority (only for the selected columns)
2.  Combine all of these datasets into a single csv file: **london_epc_certs_raw.csv**

```{r eval=FALSE, include=FALSE}
columns <- c("POSTCODE", "BUILDING_REFERENCE_NUMBER", "CURRENT_ENERGY_RATING",
  "POTENTIAL_ENERGY_RATING", "CURRENT_ENERGY_EFFICIENCY",
  "POTENTIAL_ENERGY_EFFICIENCY", "PROPERTY_TYPE", "BUILT_FORM",
  "INSPECTION_DATE", "LOCAL_AUTHORITY", "CONSTITUENCY",
  "TRANSACTION_TYPE", "ENVIRONMENT_IMPACT_CURRENT",
  "ENVIRONMENT_IMPACT_POTENTIAL", "ENERGY_CONSUMPTION_CURRENT",
  "ENERGY_CONSUMPTION_POTENTIAL", "CO2_EMISSIONS_CURRENT",
  "CO2_EMISS_CURR_PER_FLOOR_AREA", "CO2_EMISSIONS_POTENTIAL", "TOTAL_FLOOR_AREA",
  "CONSTRUCTION_AGE_BAND", "TENURE")

c_barnet <- fread(here::here("large_data/Certs_by_LA/certificates_barnet.csv"), select = columns)
c_barking_dagenham <- fread(here::here("large_data/Certs_by_LA/certificates_barking_dagenham.csv"), select = columns)
c_bexley            <- fread(here::here("large_data/Certs_by_LA/certificates_bexley.csv"), select = columns)
c_brent             <- fread(here::here("large_data/Certs_by_LA/certificates_brent.csv"), select = columns)
c_bromley           <- fread(here::here("large_data/Certs_by_LA/certificates_bromley.csv"), select = columns)
c_camden            <- fread(here::here("large_data/Certs_by_LA/certificates_camden.csv"), select = columns)
c_croydon           <- fread(here::here("large_data/Certs_by_LA/certificates_croydon.csv"), select = columns)
c_ealing            <- fread(here::here("large_data/Certs_by_LA/certificates_ealing.csv"), select = columns)
c_enfield           <- fread(here::here("large_data/Certs_by_LA/certificates_enfield.csv"), select = columns)
c_greenwich         <- fread(here::here("large_data/Certs_by_LA/certificates_greenwich.csv"), select = columns)
c_hackney           <- fread(here::here("large_data/Certs_by_LA/certificates_hackney.csv"), select = columns)
c_hammersmith_fulham <- fread(here::here("large_data/Certs_by_LA/certificates_hammersmith_fulham.csv"), select = columns)
c_haringey          <- fread(here::here("large_data/Certs_by_LA/certificates_haringey.csv"), select = columns)
c_harrow            <- fread(here::here("large_data/Certs_by_LA/certificates_harrow.csv"), select = columns)
c_havering          <- fread(here::here("large_data/Certs_by_LA/certificates_havering.csv"), select = columns)
c_hillingdon        <- fread(here::here("large_data/Certs_by_LA/certificates_hillingdon.csv"), select = columns)
c_hounslow          <- fread(here::here("large_data/Certs_by_LA/certificates_hounslow.csv"), select = columns)
c_islington         <- fread(here::here("large_data/Certs_by_LA/certificates_islington.csv"), select = columns)
c_kensington_chelsea <- fread(here::here("large_data/Certs_by_LA/certificates_kensington_chelsea.csv"), select = columns)
c_kingston          <- fread(here::here("large_data/Certs_by_LA/certificates_kingston_upon_thames.csv"), select = columns)
c_lambeth           <- fread(here::here("large_data/Certs_by_LA/certificates_lambeth.csv"), select = columns)
c_lewisham          <- fread(here::here("large_data/Certs_by_LA/certificates_lewisham.csv"), select = columns)
c_merton            <- fread(here::here("large_data/Certs_by_LA/certificates_merton.csv"), select = columns)
c_newham            <- fread(here::here("large_data/Certs_by_LA/certificates_newham.csv"), select = columns)
c_redbridge         <- fread(here::here("large_data/Certs_by_LA/certificates_redbridge.csv"), select = columns)
c_richmond          <- fread(here::here("large_data/Certs_by_LA/certificates_richmond_upon_thames.csv"), select = columns)
c_southwark         <- fread(here::here("large_data/Certs_by_LA/certificates_southwark.csv"), select = columns)
c_sutton            <- fread(here::here("large_data/Certs_by_LA/certificates_sutton.csv"), select = columns)
c_tower_hamlets     <- fread(here::here("large_data/Certs_by_LA/certificates_tower_hamlets.csv"), select = columns)
c_waltham_forest    <- fread(here::here("large_data/Certs_by_LA/certificates_waltham_forest.csv"), select = columns)
c_wandsworth        <- fread(here::here("large_data/Certs_by_LA/certificates_wandsworth.csv"), select = columns)
c_westminster       <- fread(here::here("large_data/Certs_by_LA/certificates_westminster.csv"), select = columns)

# list of all local authority certificate objects in the environment
la_list <- mget(ls(pattern = "^c_"))

# combine into a single dataframe
london_epc_certs_raw <- data.table::rbindlist(la_list, use.names = TRUE, fill = TRUE)

write.csv(london_epc_certs_raw, here::here("large_data/london_epc_certs_raw.csv"), row.names = FALSE)
```

### 1.1.1 EPCs in London

```{r}
london_epc_certs_raw <- read.csv(here::here("large_data/london_epc_certs_raw.csv"),  na.strings = c("unknown", "NO DATA!", "NODATA!", "", "INVALID!"))
```

#### Data cleaning

-   Clean column names

-   Keep only unique inspection dates (according to the latest date)

-   Remove unessessary symbols

```{r}
london_epc_certs_clean <- london_epc_certs_raw %>%
  clean_names()%>%
  #convert inspection date
  mutate(inspection_date = as.Date(inspection_date, format = "%Y-%m-%d")) %>%
  # since buildings will often have multiple inspection dates, we choose only the first one
  group_by(building_reference_number) %>%      
  arrange(desc(inspection_date)) %>%
  slice(1) %>%  
  ungroup()

# remove unnecessary symbols in construction age band
london_epc_certs_clean <- london_epc_certs_clean %>%
  mutate(construction_age_band = str_remove(construction_age_band, ".*: "))
```

-   Check how many NA values there are for each variable

```{r}
NA_val <- sapply(london_epc_certs_clean, function(x) {
  c(
    NA_count = sum(is.na(x)),
    Total = length(x),
    Percent_NA = round(sum(is.na(x))/length(x) * 100, 2)
  )})
```

-   Reassign construction_age_band values
-   Calculate median for each age band

```{r}
# check to see what age bands there are for construction age
age_bands <- london_epc_certs_clean %>%
  group_by(construction_age_band) %>%
  summarise(count = n(), .groups = "drop")

# re-assigning each construction age band
london_epc_certs_clean <- london_epc_certs_clean %>%
  mutate(construction_age_band = case_when(
    construction_age_band < "1900" ~ "pre 1900",          # anything before 1900
    construction_age_band >= "1900" & construction_age_band <= "1929" ~ "1900-1929",
    construction_age_band >= "1930" & construction_age_band <= "1949" ~ "1930-1949",
    construction_age_band >= "1950" & construction_age_band <= "1966" ~ "1950-1966",
    construction_age_band >= "1967" & construction_age_band <= "1975" ~ "1967-1975",
    construction_age_band >= "1976" & construction_age_band <= "1982" ~ "1976-1982",
    construction_age_band >= "1983" & construction_age_band <= "1990" ~ "1983-1990",
    construction_age_band >= "1991" & construction_age_band <= "1995" ~ "1991-1995",
    construction_age_band >= "1996" & construction_age_band <= "2002" ~ "1996-2002",
    construction_age_band >= "2003" & construction_age_band <= "2006" ~ "2003-2006",
    construction_age_band >= "2007" & construction_age_band <= "2011" ~ "2007-2011",
    construction_age_band >= "2012" & construction_age_band <= "2021" ~ "2012-2021",
    construction_age_band >= "2022" & construction_age_band <= "2025" ~ "2022-2025",
    construction_age_band > "2025" ~ NA,
    TRUE ~ construction_age_band  # keep everything else unchanged
  ))

london_epc_certs_clean <- london_epc_certs_clean %>%
  mutate(construction_age_band_median = case_when(
    construction_age_band < "1900" ~ NA,          # anything before 1900
    construction_age_band == "1900-1929"   ~ 1914.5,
    construction_age_band == "1930-1949"   ~ 1939.5,
    construction_age_band == "1950-1966"   ~ 1958,
    construction_age_band == "1967-1975"   ~ 1971,
    construction_age_band == "1976-1982"   ~ 1979,
    construction_age_band == "1983-1990"   ~ 1986.5,
    construction_age_band == "1991-1995"   ~ 1993,
    construction_age_band == "1996-2002"   ~ 1999,
    construction_age_band == "2003-2006"   ~ 2004.5,
    construction_age_band == "2007-2011"   ~ 2009,
    construction_age_band == "2012-2021"   ~ 2016.5,
    construction_age_band == "2022-2025"   ~ 2023.5,
    TRUE ~ NA  # keep everything else unchanged
  ))
```

-   Reassign tenure category values

```{r}
london_epc_certs_clean <- london_epc_certs_clean %>%
  mutate(tenure = case_when(
    tenure == "Rented (private)" ~ "rent_private",
    tenure == "Rented (social)" ~ "rent_social",
    tenure == "Owner-occupied" ~ "owner_occupied",
    tenure == "owner-occupied" ~ "owner_occupied",
    tenure == "rental (private)" ~ "rent_private",
    tenure == "rental (social)" ~ "rent_social",
    tenure == "Not defined - use in the case of a new dwelling for which the intended tenure in not known. It is not to be used for an existing dwelling" ~ NA))
   
```

-   Calculate energy efficiency gap

```{r}
london_epc_certs_clean <- london_epc_certs_clean %>%
  mutate(energy_efficiency_gap = potential_energy_efficiency - current_energy_efficiency)
```

-   Adding a simplified built form column

```{r}
london_epc_certs_clean <- london_epc_certs_clean %>%
  mutate(built_form_simple = case_when(
    built_form == "Detached" ~ "Detached",
    built_form == "Semi-Detached" ~ "Semi-Detached",
    built_form %in% c("Mid-Terrace", "End-Terrace",
                      "Enclosed Mid-Terrace", "Enclosed End-Terrace") ~ "Terraced",
    TRUE ~ "Not Recorded"
  ))
```

-   Adding a simplified tenure column

```{r}
london_epc_certs_clean <- london_epc_certs_clean %>%
  mutate(tenure_simple = case_when(
    tenure == "owner_occupied" ~ "owner_occupied",
    tenure %in% c("rent_private", "rent_social") ~ "rent",
    TRUE ~ "Not Recorded"
  ))
```

-   Adding a simplified property type column

```{r}
london_epc_certs_clean <- london_epc_certs_clean %>%
  mutate(property_type_simple = case_when(
    property_type == "House" ~ "House",
    property_type == "Flat" ~ "Flat",
    property_type %in% c("Maisonette", "Bungalow",
                      "Park home") ~ "Other",
    TRUE ~ "Not Recorded"
  ))
```

### Use Postcode Lookup for LA, MSOA, LSOA

```{r}
postcode_msoa_lookup <- read.csv(here::here("large_data/postcode_msoa_lookup.csv"))

london_epc_certs_clean <- london_epc_certs_clean %>%
  left_join(., postcode_msoa_lookup,
            by = c("postcode" = "pcds"))
```

-   Removing incorrect LA assignments

```{r include=FALSE}
# Check how many LAs there are (should be 32)
n_categories <- london_epc_certs_clean %>%
  summarise(n = n_distinct(ladnm, na.rm = TRUE))

unique(london_epc_certs_clean$ladnm)

# list of London LAs
london_la <- c(
  "Barking and Dagenham", "Barnet", "Bexley", "Brent", "Bromley",
  "Camden", "Croydon", "Ealing", "Enfield",
  "Greenwich", "Hackney", "Hammersmith and Fulham", "Haringey",
  "Harrow", "Havering", "Hillingdon", "Hounslow",
  "Islington", "Kensington and Chelsea", "Kingston upon Thames", "Lambeth",
  "Lewisham", "Merton", "Newham", "Redbridge",
  "Richmond upon Thames", "Southwark", "Sutton", "Tower Hamlets",
  "Waltham Forest", "Wandsworth", "Westminster"
)

# filter out rows where LA is in london list
london_epc_certs_clean <- london_epc_certs_clean %>%
  filter(ladnm %in% london_la)

# Check to see that it worked
unique(london_epc_certs_clean$ladnm)
```

FINAL london_epc_certs_clean

```{r}
write.csv(london_epc_certs_clean, here::here("large_data/london_epc_certs_clean.csv"), row.names = FALSE)
```
